---
title: "Background - ART explained"
author: "Theophanis Tsandilas and Géry Casiez"
#date: "2024-06-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We detail our derivation of the Aligned Rank Transform (ART) for the example of Figure 2. 

We use a small [dataset](ART-explained-data.csv) from a hypothetical within-subjects experiment. The dataset includes measurements $Y$ from two subjects $S$ when varying factors $A$ and $B$.  

The following code describes the construction of the aligned rank transformation for the two main effects $A$ and $B$ and the interaction $A \times B$.

```{r, echo = TRUE, message = FALSE}
library(dplyr)

# Read the dataset
df <- read.csv("ART-explained-data.csv", sep=",", header=TRUE, strip.white=TRUE)
df$S <- factor(df$S)
df$A <- factor(df$A)
df$B <- factor(df$B)

# Derive the terms for ART 
df <- df %>% 
    group_by(A) %>% mutate(Ai = mean(Y)) %>% # mean for A = Ai
    group_by(B) %>% mutate(Bj = mean(Y)) %>% # mean for B = Bj
    # mean for A=Ai, B=Bj and cell mean
    group_by(A,B) %>% mutate(AiBj = mean(Y)) %>% mutate(Cell = mean(Y)) 

df$m <- mean(df$Y) # the grand mean

# Aligned observations
alignedA <- df$Y - df$Cell + df$Ai - df$m
alignedB <- df$Y - df$Cell + df$Bj - df$m
alignedAB <- df$Y - df$Cell + df$AiBj - df$Ai - df$Bj + df$m 

# Ranks of aligned observations
df$artA <- rank(round(alignedA, digits = 7))
df$artB <- rank(round(alignedB, digits = 7))
df$artAB <- rank(round(alignedAB, digits = 7))
```

We also calculate the rank transform and the inverse normal transform:
```{r, echo = TRUE, message = FALSE}
df$rnk <- rank(df$Y) # Simple rank transform
df$int <- round(qnorm((rank(df$Y) - 0.5)/length(df$Y)), 2) # Inverse normal transform
```

We present the table with all the columns:
```{r, warning=FALSE, message = FALSE}
# Additional libraries for plotting and reporting
library(kableExtra) 
kbl(df) %>% kable_paper(position = "center") %>% scroll_box(height = "170px")
``` 

The above table shows the ranks of ART (different for each effect), RNK, and INT.  
