---
title: "The illusory promise of the Aligned Rank Transform"
subtitle: "A systematic study of rank transformations"
author: 
  - name: Theophanis Tsandilas
    orcid: 0000-0002-0158-228X
    email: theophanis.tsandilas@inria.fr
    affiliations:
      - name: Université Paris-Saclay, CNRS, Inria, LISN
        country: France
  - name: Géry Casiez
    orcid: 0000-0003-1905-815X
    email: gery.casiezuniv-lille.fr
    affiliations:
      - name: Univ. Lille, CNRS, Inria, Centrale Lille, UMR 9189 CRIStAL
        country: France
bibliography: bibliography.bib

tbl-cap-location: top
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Useful libraries
library(crosstalk)
library(kableExtra)
library(gridExtra)
library(lmerTest)
library(tidyverse)
library(plotly)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Source code for reading data from experimental results and for plotting
source("dataReaders.R")
source("plotting.R")
source("effects_plot.R")
source("interactions_plot.R")
```

```{=html}
<style>
.math.inline .MathJax  {
  font-size: 105% !important;
}

.g::before {
  content: "[G: ";
}

.g::after {
  content: "]";
}
.g {
  color: #F28C28;
}

.f {
  color: #d62b1c;
}
</style>
```

::: {.callout-note appearance="simple" icon=false collapse=false}


{{< include text/10-abstract.qmd >}}

{{< include text/20-intro.qmd >}}

{{< include text/30-background.qmd >}}

{{< include text/40-interpretingEffects.qmd >}}


## Results {#results}
Each experiment concentrated on a unique combination of distributions, experimental designs, and evaluation measures. Consequently, we organize our results into several subsections, addressing both main and interaction effects.

{{< include text/60-Results-RatioScales.qmd >}}

{{< include text/61-Results-OrdinalScales.qmd >}}

{{< include text/62-Results-AcrossExpDesigns.qmd >}}

{{< include text/63-Results-UnequalVariance.qmd >}}

{{< include text/64-Results-StatPower.qmd >}}

{{< include text/65-Results-EffectSize.qmd >}}

{{< include text/70-CaseStudies.qmd >}}

{{< include text/80-Recommandations.qmd >}}

{{< include text/90-Conclusions.qmd >}} 


## Acknowledgements {.unnumbered}
We express our gratitude to Pierre Dragicevic for introducing us to the concept of removable interactions [@Loftus:1978;@Wagenmakers:2012]. Additionally, we extend our thanks to Haiko Lüpsen for his invaluable assistance in correctly using his implementation of the multifactorial generalizations of the van der Waerden test, and the Kruskal-Wallis and Friedman tests. Finally, we are grateful to Sophie Siestrup and Daniel Martin for their responsiveness and for generously sharing their data with us. 

## References {.unnumbered}

::: {#refs}
:::
