---
title: "Appendix. Additional experimental results"
author: 
  - name: Theophanis Tsandilas
    orcid: 0000-0002-0158-228X
    email: theophanis.tsandilas@inria.fr
    affiliations:
      - name: Université Paris-Saclay, CNRS, Inria, LISN
        country: France
  - name: Géry Casiez
    orcid: 0000-0003-1905-815X
    email: gery.casiezuniv-lille.fr
    affiliations:
      - name: Univ. Lille, CNRS, Inria, Centrale Lille, UMR 9189 CRIStAL
        country: France
bibliography: bibliography.bib

tbl-cap-location: top
---

## vdWaerden

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("dataReaders.R")
source("plotting.R")

library(dplyr)
library(tidyverse)
source("plotlying.R")
```

**Type I error rates**. 

::: {#fig-vdWaerden-designs-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Appendix_test-vdWaerden"
alpha = 0.05

distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
methods = c("RNK", "INT", "VDW", "KWF")
palette = c("#E69F00", "#009E73", "#E42256", "#888888")

df <- readData(prefix, n = 20, alpha, effectType = 1, distributions, methods = methods)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 7.1, nticks=8, cbPalette = palette)
```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$)
:::

::: {#fig-vdWaerden-designs-interaction-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 7.1, nticks=8, cbPalette = palette)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$) 
:::

::: {#fig-vdWaerden-designs-interaction-2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")

df <- readData(prefix, n = 20, alpha, effectType = 0, distributions, methods = methods)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyErrorByDesign2(df, xlab = "magnitude of main effects", var = "rateX1X2", xvar = "effectX1", max = 104, nticks=8, cbPalette = palette)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$  ($n = 20$)
:::


**Power**. 

::: {#fig-vdWaerden-power-main-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
alpha <- .05

prefix <- "Appendix_test-vdWaerden-Power"
distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert5B")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")

df <- readData(prefix, n = 20, alpha, effectType = 3, distributions, methods = methods)

df <- df %>% arrange(design,distr,effectX1,rateX1)  %>% group_by(design,distr,effectX1) %>% mutate(rank = rank(rateX1))
df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyPowerByDesign(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX1", xvar = "effectX1", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
```
Ranking of methods for their power ($\alpha = .05$) to detect the **main effect of $X_1$** as a function of the magnitude of effect $a_1$ ($n=20$). Hover over the charts to inspect their power ($\%$).
:::

::: {#fig-vdWaerden-power-main-1B}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
alpha <- .05
prefix <- "Appendix_test-vdWaerden-Power-Multieffect"

df <- readData(prefix, n = 20, alpha, effectType = 3, distributions, methods = methods)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1", xvar = "effectX2", max = 104, nticks=6, cbPalette = palette)
```
Power ($\alpha = .05$) to detect the **main effect of $X_1$** ($a_1 = 0.8$) as a function of the magnitude of effect $a_2$ of the main effect of $X_2$ ($n=20$).
:::


::: {#fig-vdWaerden-power-main-2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Appendix_test-vdWaerden-Power"
df <- readData(prefix, n = 20, alpha, effectType = 4, distributions, methods = methods) 

df <- df %>% arrange(design,distr,effectX2,rateX2)  %>% group_by(design,distr,effectX2) %>% mutate(rank = rank(rateX2))
df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyPowerByDesign(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX2", xvar = "effectX2", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
```
Ranking of methods for their power ($\alpha = .05$) to detect the **main effect of $X_2$** as a function of the magnitude of effect $a_2$ ($n=20$). Hover over the charts to inspect their power ($\%$).
:::

::: {#fig-vdWaerden-power-interaction}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- readData(prefix, n = 20, alpha, effectType = 5, distributions, methods = methods)

df <- df %>% arrange(design,distr,effectX1X2,rateX1X2)  %>% group_by(design,distr,effectX1X2) %>% mutate(rank = rank(rateX1X2))
df <- as.data.frame(df) %>% reshapeByDesign(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotlyPowerByDesign(df, xlab = "magnitude of interaction effect", var = "rank", hovervar = "rateX1X2", xvar = "effectX1X2", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
```
Ranking of methods for their power ($\alpha = .05$) to detect the **interaction effect $X_1 \times X_2$** as a function of the magnitude of effect $a_{12}$ ($n=20$). Hover over the charts to inspect their power ($\%$).
:::

::: {#fig-vdWaerden-power-interaction-B}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
alpha <- .05
prefix <- "Appendix_test-vdWaerden-Power-Multieffect"

df <- readData(prefix, n = 20, alpha, effectType = 5, distributions, methods = methods)
df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

plotlyErrorByDesign2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX2", max = 104, nticks=6, cbPalette = palette)
```
Power ($\alpha = .05$) to detect the **interaction effect** (($a_{12} = 1.5$) ) as a function of the magnitude of effect $a_2$ of the main effect of $X_2$ ($n=20$).
:::

