---
title: "Introduction - Generation of the dataset for the illustrative example"
author: "Theophanis Tsandilas and Géry Casiez"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For transparency purposes, we present the simulation parameters used to generate the dataset of the introduction's illustrative example. 

**We emphasize:** this example was chosen to demonstrate how the inflated Type I error rates of ART can manifest in practice. In this scenario, ART yields very low *p-*values, whereas other methods (including the ground-truth methods) show no evidence for the corresponding effects. However, in other scenarios (and with different random seeds), ART may produce correct conclusions or exhibit less pronounced problems.

### Preliminary
Let us first import all required libraries and code we use. 

```{r, echo = TRUE, message = FALSE, warning=FALSE}
# This file includes functions for generating experimental datasets for target distributions 
source("illustrative-example-utils.R")

library(ggplot2) 
library(kableExtra)
library(gridExtra)
library(tidyverse)

```

### Setting simulation parameters
We first set the parameters of the experimental design: $4 \times 3$ within-subjects with 24 participants. The first factor (X1) represents the task *Difficulty*, while the second factor (X2) represents the *Technique*. 
```{r, warning=FALSE}
# We simulate a design with two factors. The first (X1: Difficulty) has 4 levels, and the second (X2: Technique) has 3 levels.
nlevels = c(4, 3) 

# Both factors are within-subject (repeated-measures)
within = c(1, 1)

# 24 participants
n = 24 
``` 

We then set the distribution parameters for Time measure. We use a log-normal distribution with parameters $\sigma = 0.6$ (```sigma_e``` in our code) and a target mean for responses $m = 1.4$. We set the between-participant variability to $\sigma_s = 0.4$, and the magnitude of effects to $a_1 = 2$, $a_2=0$, and $a_{12}=0$. This means that there is only an effect of *Difficulty* (first factor). For an explanation of these parameters, see our article's [method section](https://www.journalovi.org/2024-tsandilas-ranktransforms/#methodology).

```{r, warning=FALSE}
# Distribution parameters for the Time measure (log-normal)
params_lnorm  <- list(sigma_s = 0.4, sigma_e = 0.6, mean_target = 1.4)
# Setting the magnitude of effects. There is only an effect on X1: Difficulty.
coeffs_lnorm <- c("X1"=2, "X2"=0, "X1:X2"=0)

# Set the random seed
set.seed(415)
# And simulate a log-normal distribution
df_lnorm <- simulate_response(
  nlevels, within, n, coeffs_lnorm,
  family = "lnorm",
  params = params_lnorm
)

# Let's only keep two decimal digits
df_lnorm$Y = round(df_lnorm$Y, digits = 2)
``` 

Similarly, we set the distribution parameters of the *Perceived Efficiency* measure: ordinal with 5 levels. It is derived from a normal distribution with standard deviation $\sigma=1$ (```sigma_e``` in our code) and an overall mean of $m = 2.5$. We set the between-participant variability to $\sigma_s = 0.5$, while all effects are null. 

```{r, warning=FALSE}
# Distribution parameters for the measure Perceived Efficiency (ordinal with 5 levels)
params_ordinal <- list(sigma_s = 0.5, sigma_e = 1, thresholds = c(0.1, 0.4, 1, 2), mean_target = 2.5)

# Setting the magnitude of effects. All effects are null.
coeffs_ordinal <- c("X1"=0, "X2"=0, "X1:X2"=0) 

# Set the random seed
set.seed(230)
# And simulate an ordinal distribution (single Likert item)
df_ordinal <- simulate_response(
  nlevels, within, n, coeffs_ordinal,
  family = "likert",
  params = params_ordinal
)
``` 

### Formatting the dataset

Finally, we merge the two data frames into one and rename the corresponding data variables. 

```{r, warning=FALSE}
# Create the experimental data with the appropriate column names and values
labels = c("inefficient", "somewhat inefficient", "neutral", "somewhat efficient", "efficient")
df <- data.frame(
  Participant = df_lnorm$subject, 
  Difficulty = recode(df_lnorm$X1, "A1" = "Level1", "A2" = "Level2","A3" = "Level3", "A4" = "Level4"),
  Technique = recode(df_lnorm$X2, "B1" = "A", "B2" = "B","B3" = "C"),
  Time = df_lnorm$Y,
  Perceived_Efficiency = factor(df_ordinal$Y, levels = 1:5, labels = labels, ordered = TRUE)
)
``` 

We present our dataset with a table:

```{r, warning=FALSE}
kbl(df, booktabs = TRUE) %>% kable_styling(font_size=14) %>% kable_paper(position = "center") %>% scroll_box(height = "170px")
``` 

### Plotting

We also plot our data to observe the main trends.

```{r, warning=FALSE, fig.height=2.7, fig.width=8}
# Display the data
cbPalette <- c("#999999", "#E69F00", "#F15854")

p1 <- (df %>%  group_by(Participant,Technique) %>% 
summarise(Time = mean(Time), .groups="drop") %>%   
        ggplot(aes(x = Technique, y = Time, fill = Technique)) + 
        geom_boxplot(outlier.shape = NA) +
        ylab("Time (minutes) - points are means") +
        ylim(0, 8) + 
        geom_jitter(shape=20, position=position_jitter(0.1)) +
        theme_bw() + theme(legend.position = "none")) + scale_fill_manual(values=cbPalette)

p2 <- (df %>%  group_by(Participant,Difficulty,Technique) %>% 
        summarise(Time = mean(Time), .groups="drop") %>%   
        ggplot(aes(x = Difficulty, y = Time, fill = Technique)) + 
        geom_boxplot(outlier.shape = NA) +
        ylab("Time (minutes)") +
        ylim(0, 15) +
        geom_point(size = 0.7, position=position_jitterdodge(0.1)) +
        theme_bw() + theme(legend.position = "none")) +
        scale_fill_manual(values=cbPalette)

grid.arrange(p1, p2, nrow = 1, widths = c(1.5, 2.5))
``` 


```{r, warning=FALSE, fig.height=3.2, fig.width=8}
dd1 <- df %>%  group_by(Participant,Technique) %>% summarise(Perceived_Efficiency = mean(as.integer(Perceived_Efficiency)), .groups="drop")

p3 <- (dd1 %>%   
        ggplot(aes(x = Technique, y = Perceived_Efficiency, fill = Technique)) + 
        geom_boxplot(outlier.shape = NA) +
        ylab("Perceived Efficiency - circles are means") +
        scale_y_continuous(breaks = 1:5, limits = c(1,5), labels=levels(df$Perceived_Efficiency)) +
        geom_jitter(shape=1, position=position_jitter(0.2,0), size = 2) +
        theme_bw() + theme(legend.position = "none")) + scale_fill_manual(values=cbPalette) +  
        theme(axis.text.y=element_text(angle=50, hjust=0.7))

dd2 <- df %>%  group_by(Participant,Difficulty) %>% summarise(Perceived_Efficiency = mean(as.integer(Perceived_Efficiency)), .groups="drop")

diffPalette <- c("#f9f9fd", "#e2e6fa", "#bacafb", "#99b9f9")
p4 <- (dd2 %>%   
        ggplot(aes(x = Difficulty, y = Perceived_Efficiency, fill = Difficulty)) + 
        geom_boxplot(outlier.shape = NA) + ylim(1,5)+
        geom_jitter(shape=1, position=position_jitter(0.2,0), size = 2) +
        theme_bw() + theme(legend.position = "none")) + scale_fill_manual(values=diffPalette) +  
        theme(axis.title.y=element_blank(), axis.text.y=element_blank())

grid.arrange(p3, p4, nrow = 1, widths = c(2, 2))
``` 

