## Introduction {#intro}
In Human-Computer Interaction (HCI) and many areas of the behavioral sciences, researchers frequently collect data through user studies. Such data are often messy, involve small sample sizes, and violate common statistical assumptions, such as the normality assumption. To address these challenges, researchers commonly rely on nonparametric tests, which require fewer assumptions about the underlying data. However, while nonparametric tests for simple one-factorial designs are well-established, researchers face challenges in selecting appropriate methods when dealing with multifactorial designs that require testing for both main effects and interactions. 

The Aligned Rank Transform or ART [@higgins:1990; @Salter:1993; @wobbrock:2011] was introduced to address this problem by bridging the gap between nonparametric tests and ANOVA. Its popularity in HCI research has surged, in part due to the ARTool toolkit [@wobbrock:2011; @artool], which makes the method easy to apply. @Oppenlaender:2025 identify the paper by @wobbrock:2011 as a major milestone in HCI research, ranking it third in terms of citations in the ACM CHI Proceedings (1981-2024), behind Braun and Clark's [-@Braun:2006] article on thematic analysis and Hart and Staveland's [-@hart1988development] work on the NASA-TXL index. 

However, given their substantial impact on research outcomes in our community, these widely adopted methods deserve closer scrutiny. As the first author previously argued [@tsandilas:2018], the HCI community lacks sufficient expertise to critically assess the validity of statistical techniques. Our findings suggest that the widespread use of ART largely stems from limited awareness of its underlying assumptions. Many researchers treat ART as a general-purpose nonparametric alternative when ANOVA's assumptions are violated, overlooking the fact that ART’s alignment mechanism imposes even stricter assumptions --- including linear relationships between the predictors and responses, equal variances, and continuous data.

Early Monte Carlo experiments in the 1990s [@Salter:1993;@Mansouri:1995] and more recent studies [@elkin:2021] suggested that ART is a robust alternative to ANOVA when normality assumptions are violated. These results have contributed to ART's reputation as a well-established method. However, other research [@luepsen:2017; @luepsen:2018] raised concerns about the robustness of the method, demonstrating that ART fails to control the Type I error rate in many scenarios, such as when data are ordinal or are drawn from skewed distributions. Unfortunately, these warnings have received little attention, and many authors continue to rely on Wobbrock et al.'s [-@wobbrock:2011] assertion that *"The ART is for use in circumstances similar to the parametric ANOVA, except that the response variable may be continuous or ordinal, and is not required to be normally distributed."*

This paper aims to clarify the severity of these issues and understand the potential risks of the method. We present the results of a series of Monte Carlo experiments in which we simulate a broad spectrum of distributions, both discrete and continuous. Throughout the analysis, we pay particular attention to how the definition of the null hypothesis affects the interpretation of results. To improve clarity and reproducibility, we decompose the study into multiple focused experiments, each targeting a specific factor (e.g., sample size, experimental design, standard-deviation ratio) or outcome measure (Type I error rate, statistical power, and the precision of effect size estimates).

### Illustrative example 
We will begin with an illustrative example to demonstrate how the aligned rank transform can lead to false positives and a significant inflation of observed effects. This example will also serve as a brief introduction to the key concepts and methods employed throughout the paper.

Suppose a team of HCI researchers conduct an experiment to compare the performance of three user interface techniques (A, B, and C) that help users complete image editing tasks of four different difficulty levels. The experiment follows a fully balanced $4 \times 3$ within-subjects factorial design, where each participant (N = 24) performs 12 tasks in a unique order. The researchers measure the time that it takes participants to complete each task. In addition, the participants rate their perceived efficiency completing each task, using an ordinal scale with five levels: (1) *inefficient*, (2) *somewhat inefficient*, (3) *neutral*, (4) *somewhat efficient*, and (5) *efficient*.

The following table presents the experimental results: 

::: {#tbl-example}
```{r, echo=FALSE, warning=FALSE}
df <- read.csv("illustrative-example-dataset.csv", sep=",", header=TRUE, strip.white=TRUE)
kbl(df) %>% kable_paper(position = "center") %>% scroll_box(width = "740px", height = "170px")
``` 
Example dataset: Time (in minutes) spent by 24 participants and their perceived efficiency (ordinal scale) for four difficulty levels and three user interface techniques. Scroll down for full results.
:::

The experiment is hypothetical but has similarities with real-world experiments, e.g., see the experiments of @Fruchard:2023. For a detailed documentation of the simulation process we used to generate the dataset, we refer our readers to our [supplementary materials](https://github.com/journalovi/2024-tsandilas-ranktransforms/tree/main/supplementary/examples-case-studies/1-Introduction).

**Time performance.** Time measurements have been randomly sampled from a population in which: (1) *Difficulty* has a large main effect; (2) *Technique* has no main effect; and (3) there is no interaction effect between the two factors. To generate time values, we drew samples from a log-normal distribution. The log-normal distribution is often a good fit for real-world measurements that are bounded by zero and have low means but large variance [@Limbert:2001]. Task-completion times are good examples of such measurements [@Sauro:2010]. 

@fig-example presents two boxplots that visually summarize the main effects observed through the experiment. We observe that differences in the overall time performance of the three techniques are not visually clear. In contrast, time performance clearly deteriorates as task difficulty increases. 
We also observe that for the most difficult tasks (Level 4), the median time for Technique B is higher than the median time for Techniques A and B, so we may suspect that *Difficulty* interacts with *Technique*. However, since the spread of observed values is large in more difficult tasks, such differences could be the result of random noise. 

::: {#fig-example}
```{r, echo=FALSE, warning=FALSE, fig.height=3.0, fig.width=8}
cbPalette <- c("#999999", "#E69F00", "#F15854")

p1 <- (df %>%  group_by(Participant,Technique) %>% summarise(Time = median(Time), .groups="drop") %>%   
        ggplot(aes(x = Technique, y = Time, fill = Technique)) + 
        geom_boxplot(outlier.shape = NA) +
        ylab("Time (min)") +
        ylim(0, 8) + 
        geom_jitter(shape=20, position=position_jitter(0.1, 0)) +
        theme_bw() + theme(legend.position = "none") + scale_fill_manual(values=cbPalette))


p2 <- (df %>%  group_by(Participant,Difficulty,Technique) %>% summarise(Time = median(Time), .groups="drop") %>%   
        ggplot(aes(x = Difficulty, y = Time, fill = Technique)) + 
        geom_boxplot(outlier.shape = NA) +
        ylab("Time (min)") +
        ylim(0, 15) +
        #geom_jitter(shape=20, position=position_jitter(0.1)) +
        geom_point(size = 0.7, position=position_jitterdodge(0.1)) +
        theme_bw() + theme(legend.position = "none") +
        #scale_fill_brewer()
        scale_fill_manual(values=cbPalette))

grid.arrange(p1, p2, nrow = 1, widths = c(1.5, 2.5))
```
Boxplots summarizing time performance results for our illustrative example. Dots represent the mean time for each individual participant. 
:::

We opt for a multiverse analysis [@Dragicevic:2019] to analyze the data, where we conduct a repeated-measures ANOVA with four different data-transformation methods:

1. *Log transformation (LOG).* Data are transformed with the logarithmic function. For our data, this is the most appropriate method as we drew samples from a log-normal distribution. 

2. *Aligned rank transformation (ART).* Data are transformed and analyzed with the ARTool [@wobbrock:2011;@elkin:2021].

3. *Pure rank transformation (RNK).* Data are transformed with the original rank transformation [@conover:1981], which does not perform any data alignment.

4. *Inverse normal transformation (INT).* The data are transformed by using their normal scores. This rank-based method is simple to implement and has been commonly used in some disciplines. However, it has also received criticism [@Beasley:2009].  

For comparison, we also report the results of the regular parametric ANOVA with no transformation (*PAR*). To simplify our analysis and like @elkin:2021, we consider random intercepts but no random slopes. For example, we use the following R code to create the model for the log-transformed response: 

```{r, echo=TRUE, warning=FALSE}
m.log <- aov(log(Time) ~ Difficulty*Technique + Error(Participant), data = df)
```

Alternatively, we can use a linear mixed-effects model, treating the participant identifier as a random effect:

```{r, echo=TRUE, warning=FALSE}
m.log <- lmer(log(Time) ~ Difficulty*Technique + (1|Participant), data = df)
```

The table below presents the *p*-values for the main effects of the two factors and their interaction: 

|        | PAR  | LOG  | ART | RNK | INT |
|--------|------|------|-----|-----|-----|
| Difficulty  | $1.4 \times 10^{-27}$ | $2.0 \times 10^{-57}$  |  $4.5 \times 10^{-50}$ | $4.5 \times 10^{-55}$ | $3.1 \times 10^{-56}$ |
| Technique   | $.18$ | $.73$  | $.00014$ | $.61$ | $.77$ |
| Difficulty $\times$ Technique | $.26$ | $.60$ | $.00026$ | $.83$ | $.60$ |
: *p*-values for main and interaction effects {.sm}

The disparity in findings between ART and the three alternative transformation methods is striking. ART suggests that there is strong statistical evidence for all three effects. What adds to the intrigue is the fact that ART's *p*-values for *Technique* and its interaction with *Difficulty* are orders of magnitude lower than the *p*-values obtained from all other methods. We will observe similar discrepancies if we conduct contrast tests with the ART procedure [@elkin:2021], though we leave this as an exercise for the reader.

We also examine effect size measures, which are commonly reported in scientific papers. The table below presents results for partial $\eta^2$, which describes the ratio of variance explained by a variable or an interaction: 

|        | PAR  | LOG  | ART | RNK | INT |
|--------|------|------|-----|-----|-----|
| Difficulty  | $.40\ [.32, 1.0]$  | $.65\ [.60, 1.0]$  |  $.60\ [.54, 1.0]$ | $.63\ [.58, 1.0]$ | $.64\ [.59, 1.0]$  |
| Technique   | $.01\ [.00, 1.0]$ | $.003\ [.00, 1.0]$  | $.07\ [.02, 1.0]$ | $.004\ [.00, 1.0]$ | $.002\ [.00, 1.0]$ |
| Difficulty $\times$ Technique | $.03\ [.00, 1.0]$ | $.02\ [.00, 1.0]$  | $.10\ [.03, 1.0]$| $.01\ [.00, 1.0]$| $.02\ [.00, 1.0]$ |
: partial $\eta^2$ and its 95\% confidence interval {.sm}

We observe that ART exaggerates both the effect of *Technique* and its interaction with *Difficulty*. 

**Perceived efficiency.**  The perceived efficiency ratings were randomly sampled from a population with no main effect on any of the two factors and no interaction effect. @fig-example-ordinal shows that participants' ratings are concentrated at the higher end of the scale but with no clear trends. 

::: {#fig-example-ordinal}
```{r, echo=FALSE, warning=FALSE, fig.height=2.7, fig.width=8}
df$Perceived_Efficiency <- ordered(df$Perceived_Efficiency, levels = c("inefficient", "somewhat inefficient", "neutral", "somewhat efficient", "efficient"))

p3 <- (df %>%  group_by(Participant,Technique) %>% summarise(Perceived_Efficiency = mean(as.integer(Perceived_Efficiency)), .groups="drop") %>%   
        ggplot(aes(x = Technique, y = Perceived_Efficiency, fill = Technique)) + 
        geom_boxplot(outlier.shape = NA) +
        ylab("Perceived Efficiency") +
        scale_y_continuous(breaks = 1:5, limits = c(1,5), labels=levels(df$Perceived_Efficiency)) +
        geom_jitter(shape=1, position=position_jitter(0.2,0), size = 2) +
        theme_bw() + theme(legend.position = "none")) + scale_fill_manual(values=cbPalette) +  
        theme(axis.text.y=element_text(angle=50, hjust=0.7))

#diffPalette <- c("#f9f9f9", "#e2e2e2", "#b9b9b9", "#999999")
diffPalette <- c("#f9f9fd", "#e2e6fa", "#bacafb", "#99b9f9")
p4 <- (df %>%  group_by(Participant,Difficulty) %>% summarise(Perceived_Efficiency = mean(as.integer(Perceived_Efficiency)), .groups="drop") %>%   
        ggplot(aes(x = Difficulty, y = Perceived_Efficiency, fill = Difficulty)) + 
        geom_boxplot(outlier.shape = NA) + ylim(1,5)+
        geom_jitter(shape=1, position=position_jitter(0.2,0), size = 2) +
        theme_bw() + theme(legend.position = "none")) + scale_fill_manual(values=diffPalette) +  
        theme(axis.title.y=element_blank(), axis.text.y=element_blank())

grid.arrange(p3, p4, nrow = 1, widths = c(3, 3))
```
Boxplots summarizing the results on efficiency as perceived by participants. Circles represent the mean rating of each individual participant. 
:::

For our analysis, we compare the results of PAR, ART, RNK, and INT, since LOG is not relevant for this type of data. We also report the results of the Friedman test [@Friedman] for each factor, where we first aggregate the data by taking the mean over the other factor. Results are as follows:  

|        | PAR  | ART | RNK | INT | Friedman Test |
|--------|------|-----|-----|-----|-----|
| Difficulty  | $.43$ | $.00018$  |  $.53$ | $.49$ | $.85$ |
| Technique   | $.73$ | $.00013$  | $.90$ | $.82$ | $.92$ |
| Difficulty $\times$ Technique | $.56$ | $.00055$ | $.33$ | $.43$ | |
: *p*-values for main and interaction effects {.sm}

ART's *p*-values for all three effects are surprisingly low, while no other method shows any substantial evidence for such effects. We also present the methods' effect size estimates:

|        | PAR  | ART | RNK | INT |
|--------|------|-----|-----|-----|
| Difficulty  | $.01\ [.00, 1.0]$  | $.08\ [.03, 1.0]$  |  $.009\ [.00, 1.0]$ | $.009\ [.00, 1.0]$ |
| Technique   | $.003\ [.00, 1.0]$ | $.07\ [.02, 1.0]$  | $.0009\ [.00, 1.0]$ | $.002\ [.00, 1.0]$ |
| Difficulty $\times$ Technique | $.02\ [.00, 1.0]$ | $.09\ [.03, 1.0]$  | $.03\ [.00, 1.0]$| $.02\ [.00, 1.0]$|
: partial $\eta^2$ and its 95\% confidence interval {.sm}

As we demonstrate in this paper, such discrepancies stem from ART's inability to handle discrete distributions, and unfortunately, the issue becomes more pronounced with larger sample sizes. @luepsen:2017 previously warned researchers about this problem, but his warnings have largely been ignored.

### Overview
The results from the above example are not simply due to random patterns in the data, but can be explained by considering the strict assumptions of ART's alignment procedure. We show that ART’s error inflation is systematic across a range of distributions that deviate from normality, including both continuous and discrete cases.

The paper is structured as follows. [Section 2](#background) introduces nonparametric tests and rank transformations, and explains how ART is constructed. It also provides a summary of previous experimental results regarding the robustness of ART and other related rank-based procedures, along with a survey of recent studies using ART. [Section 3](#assumptions) clarifies the method's distributional assumptions. [Section 4](#interpretation) investigates issues regarding the definition of the null hypothesis and the interpretation of main and interaction effects. [Section 5](#methodology) outlines our experimental methodology, while [Section 6](#findings) presents our findings. [Section 7](#casestudies) revisits results of previous studies employing ART and  illustrates how its application can lead to erroneous conclusions. [Section 8](#recommendations) offers recommendations for researchers. Finally, [Section 9](#conclusions) concludes the paper.

In addition to the main sections of the paper, we provide an appendix (see [Appendix I](appendix_I.html)) with results from additional Monte Carlo experiments.
